---
title: "Subsetting Practice"
author: "Ziqi Liu"
date: "May 16, 2018"
output: 
  html_document:
    keep_md: true
    toc: true
    toc_float: true
    toc_depth: 2
editor_options:
  chunk_output_type: console
---

# 1.0 Subsetting
- 3 subsetting operators
- 6 types of subsetting

# 2.0 Selecting multiple elements
- use [] operator

## 2.1 Atomic vectors
- positive integers return elements at specified positions
```{r}
x <- c(2.1, 4.2, 3.3, 5.4)
x[c(3,1)]
x[order(x)]
```

- duplicated indices yield duplicated values
```{r}
x[c(1, 1)]
```

- real numbers are truncated to integers (round down)
```{r}
x[c(2.1, 2.9)]
```

- negative integers omit elements at the specified positions
```{r}
x[-c(3, 1)]
```

- you can't mix positive and negative integers in a single subset
```{r}
# x[c(-1, 2)]
```

- logical vectors select elements where the corresponding logical value is TRUE
```{r}
x[c(T, T, F, F)]
x[x>3]
```

- if the logical vector is shorter than the vector being subsetted, it will be recycled to be the same length
```{r}
x[c(T, F)]
#same as
x[c(T, F, T, F)]
```

- a missing value in the index yields a missing value in the output
```{r}
x[c(T, T, NA, F)]
```

- nothing returns the original vector/matrix/df/array
```{r}
x[]
```

- if a vector is named you can also use character vectors to return elements with matching names (must be exact match)
```{r}
y <- setNames(x, letters[1:4])
y

y[c("d","c","a")]

#like integer indices you can repeat them
y[c("a","a","a")]
```

## 2.2 Lists
- using [[]] and $ lets you pull out components of a list
- you can subset higher-dimensional structures with multiple vectors, a single vector, or a matrix
```{r}
a <- matrix(1:9, nrow = 3)
colnames(a) <- c("A", "B", "C")
a
a[c(TRUE, FALSE, TRUE), c("B", "A")]
a[0, -2]
```

## 2.3 Data frames
- dfs possess characteristics of both lists and matrices, if you subset with a single vector they behave like lists, if you subset with 2 vectors they behave like matrices
```{r}
df <- data.frame(x=1:3, y=3:1, z=letters[1:3])
df
```

```{r}
# to select rows
df[df$x == 2, ]
df[c(1, 3),]

# to select columns
# 1. like a list
df[c("x", "z")]
# 2. like a matrix
df[, c("x", "z")]
```

## 2.4 Preserving dimensionality
- subsetting 2d data structures with a single number/char/logical vector will simplify the returned output 
- to preserve original dimensionality, use drop=F
```{r}
# for matrices and arrays, any dimensions with len=1 will be dropped
a <- matrix(1:4, nrow=2)

str(a[1, ]) #1D
str(a[1, , drop=F]) #2D

# data frames with a single column will return just that column
df <- data.frame(a=1:2, b=1:2)

str(df[, "a"])
str(df[, "a", drop=F])
```

- ALWAYS ADD drop=F JUST IN CASE!!

## 2.5 Exercises

1)
```{r}
# mtcars[mtcars$cyl=4, ] will error need to use equality operator ==

mtcars[mtcars$cyl==4, ]
```

2)
```{r}
# mtcars[-1:4, ] will error because you can't index over a negative and positive integer

mtcars[2:4, ] # this will drop the first row, and then go to 4
mtcars[c(ncol(mtcars), 1:4), ] # this will get the last row and the first 4
```

3)
```{r}
# mtcars[mtcars$cyl <= 5] will error because we haven't specified rows or coluns

mtcars[mtcars$cyl <=5, ]
```

4)
```{r}
# mtcars[mtcars$cyl == 4 | 6, ] 

mtcars[mtcars$cyl == 4 | mtcars$cyl == 6, ]
```

5) 
why does x <- 1:5; x[NA] yield five missing values? (Hint: why is it different from x[NA_real_]?)

6)
What does upper.tri() return? How does subsetting a matrix with it work? Do we need any additional subsetting rules to describe its behaviour?
```{r}
x <- outer(1:5, 1:5, FUN = "*") ; x
# this creates a 5 by 5 matrix where each value is its row*col

upper.tri(x)
# creates a matrix of T or F where T is the upper triangle of values

x[upper.tri(x)]
# this subsets x with only the values that were true, so values in the upper triangle
```

7. Why does mtcars[1:20] return an error? How does it differ from the similar mtcars[1:20, ]?
```{r}
# mtcars[1:20] doesn't work because we didn't specify rows or cols
mtcars[1:20, ] #returns rows 1 to 20
```

8. Implement your own function that extracts the diagonal entries from a matrix (it should behave like diag(x) where x is a matrix).
```{r}
x <- matrix(LETTERS[1:25], ncol=5) ; x
diag(x)
```

9. What does df[is.na(df)] <- 0 do? How does it work?
```{r}
df <- data.frame("memes" = letters[1:9], "dreams" = 7:15, "123" = c(1, 69, NA, 23, 61, 2, 36, NA, 3045069)) ; df
# creates df with some NA

is.na(df)
# creates a matrix of logicals where T is NA

df[is.na(df)] <- 0 ; df
# subsets df looking for places where the value is NA and then turns that to 0
```

# 3.0 Selecting a single element
- use $ or [[]]
- [[]] is like a nested list
```{r}
mtcars[1] #returns the first column with all the rownames
mtcars[[1]] #returns all the values of the first column
```

# 3.1 $
- x$y is ~= x[["y"]]
- where x is the tibble/df/matrix name and y is the name of a variable(column) in it

# 3.2 Exercises
1)
Come up with as many way as possible to extract the third value from the cyl variable in the mtcars dataset.

```{r}
mtcars[[2]][3]
mtcars$cyl[3]
```

# 4.0 Subsetting and assignment
- all subsetting operators can be combined with assignment operators to modify selected values of input vector
```{r}
x <- 1:5 ; x[c(1, 2)] <- 2:3 ; x

x[-1] <- 4:1 ; x
# length(LHS) needs to = length(RHS) for assignment to work

x[c(1, 1)] <- 2:3 ; x
# duplicated indices go unchecked and may be problematic

# you can't combine integer indices with NA but you can combine logical indices with NA
```

- subsetting with nothing and then assignment preserves original object class and structure
```{r}
mtcars[] <- lapply(mtcars, as.integer) #mtcars stays as df
mtcars <- lapply(mtcars, as.integer) #mtcars becomes list
```

# 5.0 Applications

## 5.1 Lookup tables
```{r}
x <- c("m", "f", "u", "f", "f", "m", "m")
lookup <- c(m = "Male", f = "Female", u = NA)
lookup[x]
unname(lookup[x])
```

## 5.2 Matching or merging by hand (integer subsetting)

```{r}
grades <- c(1, 2, 2, 3, 1)
info <- data.frame(
  grade = 3:1,
  desc = c("Excellent", "Good", "Poor"),
  fail = c(F, F, T)
)
id <- match(grades, info$grade)
info[id, ]
```

## 5.3 Random samples/bootstrap (integer subsetting)

- you can use integer indices to perform random sampling or bootstrapping
```{r}
df <- data.frame(x = rep(1:3, each = 2), y = 6:1, z = letters[1:6]) ; df
df[sample(nrow(df)), ] #reorder rows
df[sample(nrow(df), 3), ] #select 3 random rows
```

## 5.4 Ordering (integer subsetting)

- order() takes a vector as input and returns an integer vector describing how the vector should be ordered
```{r}
x <- c("b", "c", "a")
order(x)
x[order(x)]
```

- you can order in desc order by using decreasing = T
``` {r}
df
df2 <- df[sample(nrow(df)), 3:1] #randomly reorder df rows, invert columns
df2[order(df2$x), ] #order values of col x
df2[, order(names(df2))] #order by col names
```

## 5.5 Expanding aggregated counts (integer subsetting)

- sometimes you get a data frame where identical rows have been collapsed into one and a count column has been added
- rep() and integer subsetting make it easy to uncollapse the data by subsetting with a repeated row index
```{r}
df <- data.frame(x=c(2,4,1), y=c(9,11,6), n=c(3,5,1)) ; df
rep(1:nrow(df), df$n)
df[rep(1:nrow(df), df$n), ]
```

## 5.6 Removing columns from data frames (character subsetting)

- 2 ways to remove columns from df
- you can set columns to NULL
```{r}
mtcars$cyl <- NULL ; head(mtcars)
```

- or you can subset to return only the columns you want
```{r}
rm(mtcars)
mtcars[-2]
```

## 5.7 Selecting rows based on a condition (logical subsetting)

```{r}
mtcars[mtcars$gear == 5, ]
mtcars[mtcars$gear == 5 & mtcars$cyl == 4, ]

# subset function also exists! these are equivalent to:

subset(mtcars, gear ==5)
subset(mtcars, gear == 5 & cyl == 4)
```

## 5.8 Boolean algebra vs sets (logical & integer subsetting)

- which() allows you to convert a boolean representation to an integer representation
```{r}
x <- sample(10) < 4 ; x
which(x)

(x1 <- 1:10 %% 2 == 0)
(y1 <- 1:10 %% 5 == 0)
(x2 <- which(x1))
(y2 <- which(y1))

# X & Y <-> intersect(x,y)
x1 & y1
intersect(x2, y2)

# X | Y <-> union(x,y)
x1 | y1
union(x2, y2)

# X & !Y <-> setdiff(x,y)
x1 & !y1
setdiff(x2, y2)
```

## 5.9 Exercises
3)
How could you put the columns in a data frame in alphabetical order?
```{r}
mtcars[ , order(names(mtcars))]
```



